---
title: "Coming up in the next version"
description: |
  This article gives a preview of what is to come in the next mrgsolve release.
author:
  - name: Kyle Baron
    url: https://github.com/kylebaron
    affiliation: Team mrgsolve
    affiliation_url: https://mrgsolve.org
date: "`r Sys.Date()`"
categories:
  - what's new
output: 
  distill::distill_article: 
    toc: true
    toc_float: true
    number_sections: true
    highlight: pygments
    highlight_downlit: false 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mrgsolve)
library(tidyverse)
```

There are a bunch of new features coming up in the next mrgsolve release. Most 
of these are related to model syntax (what / how you write your model in the 
`.mod` or `.cpp` file) rather than model functionality. 

# Use THETA(n) to refer to THETAn

All models will be able to use `THETA(n)` to refer to `THETAn`. For example:


```{c, eval = FALSE}
[ nmxml ] 
run = 100
root = "cppfile"

[ main ] 
double CL = THETA(1) * pow(WT/70, 0.75);
```

You'll notice that we referred to a nonmem run with `[ nmxml ]` so this syntax
is meant be one small step toward making it easier to translate your model 
from the nonmem control stream to mrgsolve format. This is pretty simple, but 
will mean less time removing parentheses when copying over model code. 

# New plugin called `autodec`

Plugins are extra functionality that you can invoke when coding your model. 
For example, plugins exist for calculating time after dose, adding Rcpp 
functionality to your model etc and [more](https://mrgsolve.org/user_guide/model-specification.html#block-plugin).

One new plugin is called `autodec`. This plugin will automatically declare 
C++ variables for you. For example, I can code the snipped from the previous
block using `autodec` plugin and mrgsolve will discover that `CL` needs to be
declared:

```{c, eval = FALSE}
[ plugin ] autodec

[ nmxml ] 
run = 100
root = "cppfile"

[ main ] 
CL = THETA(1) * pow(WT/70, 0.75);
```

The motivation behind this feature is to make it easier to code models with 
lots of variables when all your variables are simple double precision floating 
point numbers.

# New plugin called `nm-vars`

This plugin provided a nonmem-like syntax for _certain model elements_, 
including 

- `Fn`, `Dn`, `Rn`, and `ALAGn`
- `A(n)`, `A_0(n)` and `DADT(n)`

So your mrgsolve model translated from nonmem might look like this

```{r, results = "hide"}
see(modlib("1005"))
```

```{c, eval=FALSE, code = modlib("1005", compile = FALSE)@code}

```

where I updated the internal model `1005` to reflect use of both `autodec` and 
`nm-vars` plugins. 

Notice that __not all nonmem syntax is supported__; just select data structures
that use `A` and `DADT` to refer to compartments and differential equations.
Also notice that we still require `;` at the end of each line and we still 
need to use `pow(base, exponent)` to calculate exponent of some number. 

I'll have full documentation in the user guide once the release is made. 


# New time spacer for event sequences

The new feature is an `ii` spacer for sequences of event objects. 

Using [event objects](https://mrgsolve.org/user_guide/event-chapter.html), 
we can create complicated dosing regimens from simpler ones. For example, we
might have a week of 50 mg dosing followed by a week of 100 mg dosing and then 
up to 200 mg dosing. We can accomplish that with an event sequence

First create the individual pieces

```{r}
a <- ev(amt = 50,  ii = 24, addl = 6)
b <- ev(amt = 100, ii = 24, addl = 6)
c <- ev(amt = 200, ii = 24, addl = 6)
```

Then put them in a sequence

```{r}
data <- ev_seq(a, b, c)
data
```

When they are sequenced, you'll see that the second piece (`b`) starts one
dosing interval after the last dose in the first piece (`a`). 

We can put a 24 hour spacer between `a` and `b`

```{r}
seq(a, wait = 24, b)
```

Here, the last dose in `a` is given, we wait one dosing interval (24 hours) then 
wait another 24 hours (via `wait`) and then start `b`. 

Rather than using `wait`, we can use `ii` to specify the amount of time 
from the last dose in `a` to the first dose in `b`. So if we want to wait 3 days
between the last dose in `a` and the first dose in `b`

```{r}
ev_seq(a, ii = 3*24, b)
```

Notice that this same behavior can be achieved with `wait` but using `ii` 
might be easier to interpret in some cases. 

#  Collapse matrices

Sometimes in your model, you have multiple OMEGA or SIGMA blocks like this

```{r}
code <- '
$OMEGA  @name first
1 2 3

$OMEGA @name second
4
'
mod <- mcode("collapse", code, compile = FALSE)

omat(mod)
```


If I want to update these values but I only have a 4x4 matrix

```{r, include = FALSE}
mat <- as.matrix(omat(mod))  / 10

```

```{r}
mat
```

I can't do it; mrgsolve wants a 3x3 matrix for the first slot and 1x1 matrix
for the second. 

A new function will collapse the model matrix into larger matrices

```{r}
mod <- collapse_omega(mod)
omat(mod)
```

```{r}
mod <- update(mod, omat = mat)
omat(mod)
```

You can't split matrices up into smaller chunks, but if there are more than 
two blocks, you can select which blocks to join

```{r}
code <- '
$OMEGA  @name first
1 2 3

$OMEGA @name second
4

$OMEGA @name third
5 6 7 8
'
mod <- mcode("collapse2", code, compile = FALSE)

omat(mod)

mod <- collapse_omega(mod, range = c(2, NA), name = "remainder")
omat(mod)
```


# Pull Requests

A lot of the details around these new features were hammered out on public
pull requests

- https://github.com/metrumresearchgroup/mrgsolve/pull/891
- https://github.com/metrumresearchgroup/mrgsolve/pull/892
- https://github.com/metrumresearchgroup/mrgsolve/pull/900
- https://github.com/metrumresearchgroup/mrgsolve/pull/901
- https://github.com/metrumresearchgroup/mrgsolve/pull/907

Any further feedback or perspective on these changes or any testing that you 
can do before release is always appreciated. 


